Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-20 doc/mc-paper.tex
Minor tweaks

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And/Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a d-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

In this paper, we introduce \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers, including d-DNNF\@. To capture both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF), we propose the CRAT file format. A CRAT
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that a CRAT file must satisfy in order to
encode a well-formed POG and provide a valid equivalence proof.

@@

Figure~\ref{fig:chain} illustrates our certifying knowledge compilation
and (weighted) model counting toolchain.  A
{\em proof generator} generates a CRAT description from a d-DNNF
graph produced by \dfour{}, a state-of-the-art knowledge compiler.
A {\em proof checker} then verifies that the combination of CNF and CRAT files
satisfy our rules, and a {\em ring evaluator} can compute
either a standard or weighted model count from the POG\@ representation.
We evaluate the capabilities of this toolchain using
benchmark formulas from the 2022 standard and weighted model
competitions.  Our tools are able to handle all but the largest d-DNNF
graphs generated by \dfour{}.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

@@

>>>>

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a d-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

This paper introduces \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers. To capture both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF), we propose the CRAT file format. A CRAT
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that a CRAT file must satisfy in order to
encode a well-formed POG and provide a valid equivalence proof.

@@

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler
generates a d-DNNF representation, and the {\em proof
  generator} uses this to generate a CRAT file.
The {\em proof checker} verifies the equivalence of the CNF and CRAT representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

@@

We evaluate the capabilities of this toolchain using benchmark
formulas from the 2022 standard and weighted model competitions.  Our
tools are able to handle all but the largest d-DNNF graphs generated
by \dfour{}.  We also measure the benefits of several optimizations as
well as the performance of the verified checker relative to one
written in C.
